Modal operators for a constructive account of well quasi-orders

### 2025-02-27 slides-abmv2024.tex
More notes

    For positive toposes~$\E$, a geometric implication holds in~$\E$ iff it
    holds in the base universe. For positive overt toposes~$\E$, we even have
    that a bounded first-order formula holds in~$\E$ iff it holds in the base.
    Hence, for the purpose of verifying a bounded first-order assertion about
    the base, we can freely pass to a positiv overt topos with problem-adapted
    better higher-order properties (such as that some uncountable set from the
    base now appears countable, or that an infinite sequence whose existence is
    predicted by failing dependent choice now actually exists).

>>>>

    For positive toposes~$\E$, a geometric implication holds in~$\E$ iff it
    holds in the base universe. For positive overt toposes~$\E$, we even have
    that a bounded first-order formula holds in~$\E$ iff it holds in the base.
    Hence, for the purpose of verifying a bounded first-order assertion about
    the base, we can freely pass to a positive overt topos with problem-adapted
    better higher-order properties (such as that some uncountable set from the
    base now appears countable, or that an infinite sequence whose existence is
    predicted by failing dependent choice now actually exists).

### 2025-02-27 slides-paris2025.tex
More notes

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{2.1em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
\begin{frame}{Well quasi-orders}
  \jnote{2-3}{
    The presented proof rests on the law of excluded middle and hence cannot
    immediately be interpreted as a program for finding suitable indices~$i <
    j$. However, constructive proofs are also possible (for instance by
    induction on the value of a given term of the sequence, see
    \href{https://www.math.lmu.de/~petrakis/Dickson.pdf}{Constructive
    combinatorics of Dickson's Lemma} by Iosif Petrakis for several fine
    quantitative results). And even more:
    There is a procedure for regarding this proof---and many others in the
    theory of well quasi-orders---as \emph{blueprints} for more informative
    constructive proofs. This shall be our motto for today:

@@

  \jnote{4}{
    The dependence of the theory on well quasi-orders on classical transfinite
    methods is already present in one of basic observations of this theory:

@@

    The appeal to dependent choice can be removed by always picking the
    smallest possible next index in~$\NN \setminus K$, doable by yet another
    invocation of~\bad{\textsc{lem}}, but the result remains fundamentally
    noneffective---in the special case~$X = (\{ 0, 1 \}, {=})$, the statement
    of the lemma implies the infinite pigeonhole principle.
  }

@@

    Another example:
    In the community around Krull's lemma, it is well-known that
    we can constructively infer that a given ring element~$x \in A$ is nilpotent from
    knowing that it is contained in the \emph{generic prime ideal} of~$A$. This entity
    is not actually an honest prime ideal of the ring~$A$ in the base
    universe, but a certain combinatorial notion (efficiently dealt with using
    \emph{entailment relations}). Constructive forcing allows us to reify the
    generic prime ideal as an actual prime ideal in a suitable forcing extension.
  }

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-4.68em}\includegraphics[width=1.2\paperwidth]{wqo-faded}\end{minipage}}
\begin{frame}{Multiversal constructive combinatorics}
  \justifying
  \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
  is well$\ind$.

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Multiversal constructive algebra}
  \justifying
  \textbf{Thm.}
  Let~$M$ be a surjective matrix with more rows than columns over a
  ring~$A$. Then~$1 = 0$ in~$A$.
  \medskip

@@

>>>>

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{2.1em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
\begin{frame}{Well quasi-orders}
  \jnote{1}{
    Well quasi-orders are an important notion in proof theory and termination
    analysis.
  }

  \jnote{2-3}{
    The presented proof rests on the law of excluded middle and hence cannot
    immediately be interpreted as a program for finding suitable indices~$i <
    j$. However, constructive proofs are also possible (for instance by
    induction on the value of a given term of the sequence, see
    \href{https://www.math.lmu.de/~petrakis/Dickson.pdf}{Constructive
    combinatorics of Dickson's Lemma} by Iosif Petrakis for several fine
    quantitative results). And even more:
    There is a procedure for regarding this proof---and many others in the
    theory of well quasi-orders---as \emph{blueprints} for more informative
    constructive proofs. This shall be our motto for today:

@@

  \jnote{4}{
    \vspace*{-1em}
    The dependence of the theory on well quasi-orders on classical transfinite
    methods is already present in one of the first and central observations of this theory:

@@

    The appeal to dependent choice can be removed by always picking the
    smallest possible next index in~$\NN \setminus K$, doable by yet another
    invocation of~\bad{\textsc{lem}}. But the result remains fundamentally
    noneffective---in the special case~$X = (\{ 0, 1 \}, {=})$, the statement
    of the lemma implies the infinite pigeonhole principle.
  }

@@

    Another example:
    In the community around Krull's lemma, it is well-known that
    we can constructively infer that a given ring element~$x \in A$ is nilpotent from
    knowing that it is contained in the \emph{generic prime ideal} of~$A$. This entity
    is not actually an honest prime ideal of the ring~$A$ in the base
    universe, but a certain combinatorial notion (efficiently dealt with using
    \emph{entailment relations}). Constructive forcing allows us to reify the
    generic prime ideal as an actual prime ideal in a suitable forcing extension,
    so in a suitable topos (der little Zariski topos of the ring).
  }

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-4.68em}\includegraphics[width=1.2\paperwidth]{wqo-faded}\end{minipage}}
\begin{frame}{Multiversal constructive combinatorics}
  \jnote{1}{
    The displayed multiversal proof closely mimics the classical proof (for well\sinf), but
    is fully constructive (for well\ind). It would be possible to streamline
    this proof and unroll the topos-theoretic machinery, to obtain an explicit
    algorithm of type
    \[ \mathsf{Good} \mid_X [\,] \times \mathsf{Good} \mid_Y [\,] \ \longrightarrow\
      \mathsf{Good} \mid_{X \times Y} [\,].
    \]

    The modal language was recently used to answer a question by Stefano
    Berardi, Gabriele Buriola and Peter Schuster, see
    \href{https://www.speicherleck.de/iblech/stuff/slides-abmv2024.pdf\#page=16}{this
    set of slides}.
  }

  \justifying
  \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
  is well$\ind$.

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Multiversal constructive algebra}
  \jnote{1-}{
    The displayed classical proof is quite efficient from the point of view of
    organizing mathematical knowledge, as it quickly reduces the general
    situation of dealing with an arbitrary ring to dealing with a field.
    Alas, read literally, it is hopeless ineffective.
  }

  \jnote{2-}{
    By employing modal language, we can closely mimic the original proof and be
    fully constructive at the same time.
  }

  \jnote{3-}{
    By unwinding all modal definitions, the modal proof can be unrolled to a
    fully explicit computation.
  }

  \justifying
  \textbf{Thm.}
  Let~$M$ be a surjective matrix with more rows than columns over a
  ring~$A$. Then~$1 = 0$ in~$A$.
  \medskip

@@

\begin{frame}
  \emph{Backup slides.}
\end{frame}
